Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-08-12 analysis/paper/paper.Rmd
edit discussion

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans with a high diversity of types [@Cheng2008; @Wang2011]. This hinted at the emergence of social inequality within the Indigenous community. To understand the relationship between social inequality and foreign presences, we examined ceramic vessel shape standardization to measure craft specialization as a proxy for social change [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and the pre-European and Chinese periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Such shape is more pronounced in the Chinese period. In general, vessels become oval-shaped with a restricted mouth over time, which corresponds with an increase in body diameter of vessels. The correlation between shape and body diameter suggests that body diameter significantly varies with vessel shape. Oval-shaped vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.

@@

For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods. A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting that a more standardized pottery shape after contact with the Chinese. In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period. Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period when we use body diameter as our proxy variable of size. This might hint an intentional behavior to achieve a homogeneous form of vessels, but we still need more metric variables to assess the variation in size. Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. We note that a small sample size in the Chinese period may lead to a more standardized shape. However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001]. Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].

@@

Whether standardization suggests an emergence of specialized production depends on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylish and functional attributes [@Costin2001; @Costin1995]. In other words, mechanical standardization is associated with potter's skills and experiences that increased morphological uniformity normally reflects fewer producers. The population at Kiwulan could be an indicator to estimate the potential number of producers. According to the historical records [@Nakamura1938, pp. 12], the population in Kiwulan reached a higher number in the 17th century as the European presence and declined in the Chinese period due to movement of Indigenous people to the south [@Chen2007]. The changes in population size described in documents correspond to the observation of ceramic abundance at Kiwulan. Thus, we model cv values as a function of the mass of ceramics using Poisson GLM with a link function. The model suggests that ceramic abundance strongly predicts the cv values (β = $`r beta_print`$, p = $`r pval_print`$). This indicates that a more standardized shape in the Chinese period may be influenced by a small population if ceramic abundance can reflect the population size.

However, intentional standardization could also contribute to shape standardization in our case. To ensure the standardization is caused by fewer producers or specialized organization, we investigate other factors that may contribute to shape standardization, such as the function and decoration of ceramics. To determine the function of pots in relation to shape changes, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909]. Stylish attributes present a wide variation in geometric motifs from pot to pot across three phases. In general, the ceramics in the Chinese period have slightly fewer types of motifs according to the ratio of distinct motifs to the total number for each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020]. The evidence shows that we could not completely rule out the intentional standardization in our sample.

>>>>

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods with high diversity in types, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011]. This hinted at the emergence of social inequality within the Indigenous community. To understand the relationship between social inequality and foreign presences, we examined shape standardization of ceramics to measure craft specialization as a proxy for social change [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates significant differences in shapes between the pre-European and Post-European periods, and pre-European and Chinese periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Such shape is more pronounced in the Chinese period. In general, vessels become oval-shaped with a restricted mouth over time, which corresponds with an increase in body diameter of vessels. The correlation between shape and body diameter suggests that body diameter significantly varies with vessel shape. Oval-shaped vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.

@@

For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods. A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`r pre_chi`$) or post-European periods (p= $`r post_chi`$) [@WangandMarwick2020], suggesting a more standardized shape after contact with the Chinese. In addition, we found a more homogeneous shape accompanied by a more standardized size, represented by the body diameter of vessels, in the Chinese period. Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period when we use body diameter as our proxy variable of size. This might hint an intentional behavior to achieve a homogeneous form of vessels, but we still need more metric variables to assess the variation in size. Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. We note that a small sample size in the Chinese period may lead to a more standardized shape. However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001]. Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].

@@

Whether the shape standardization we found suggests craft specialization depends on the number of producers, which distinguishes mechanical standardization from intentional standardization defined by stylish and functional attributes [@Costin2001; @Costin1995]. Mechanical standardization is used to access the appearance of specialized production based on the assumption that increased skills, routinization, and lower diversity of producers will lead to morphological uniformity [@Arnold2000]. The population at Kiwulan could be an indicator to estimate the potential number of producers. According to the Dutch census in 1648 [@Nakamura1938, pp. 12], the population at Kiwulan was large but declined in the Chinese period due to the movement of Indigenous people to the south [@Chen2007]. The change in population size corresponds with the observation of ceramic abundance at Kiwulan. Thus, we model CV values as a function of the mass of ceramics using Poisson GLM with a link function. The model suggests that ceramic abundance strongly predicts the CV values (β = $`r beta_print`$, p = $`r pval_print`$). This indicates that a more standardized shape in the Chinese period may be influenced by a small population if ceramic abundance can reflect the population size.

However, intentional standardization could also contribute to shape standardization in our case. To ensure the standardization is caused by a small number of specialized potters, we investigate other factors that may contribute to shape standardization, including the function and decoration of ceramics. To determine the function of pots in relation to shape changes, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909]. Stylish attributes present a wide variation in geometric motifs from pot to pot across three phases. In general, the ceramics in the Chinese period have slightly fewer types of motifs according to the ratio of distinct motifs to the total number for each phase (Chinese = `r chinese_deco`, post-European = `r post_deco`, pre-European = `r pre_deco`) [@WangandMarwick2020]. The evidence shows that we could not completely rule out the intentional standardization in our sample.
